Mechanized Reasoning about a Capability Machine

### 2019-10-17 prisc-2020/paper.tex
editing pass

In summary, we present Iris formalizations of 
\begin{itemize}
  \item a program logic for reasoning about capability machine programs.
  \item the logical relation from \cite{skorstengaardESOP18}, 
     which captures capability safety and which can be used to reason about WBCF and LSE.
\end{itemize}
All of the technical development is mechanized in Coq using the Coq implementation of Iris.

\todo{Alix: I think we need a short discussion (not specifically here) about the fact that we don't really have a theorem that says something like "being in the logical relation implies LSE and WBCF" as that's not really well defined, but we can prove things about specific programs that rely on it}

@@

\section{The Program Logic of a Capability Machine}

%\begin{itemize}
%\item we define the operational semantics of the CM
%\item we prove basic hoare triples for each of the atomic instructions
%  of the CM (i.e., for load, store, etc.)
%\item non-trivial
%  \begin{itemize}
%  \item while Iris is a framework and supports many languages, it is
%    geared towards models of higher-level languages which abstract from
%    program counters and the fact that programs are stored in memory.
%    -- here our low-level machine model includes registers, in particular
%    the program counter register, and instructions are stored in memory.
%  \item to capture the semantics of the CM in the Iris framework, 
%    we introduce an abstract notion of instruction, whose
%    operational meaning is to execute the next instruction the program counter
%    points to, and abstract notions of values, failed and halted.
%  \item for each machine instruction, we can then prove a hoare triple
%    which expresses the pre- and post-conditions for executing the instruction
%  \end{itemize}
%\end{itemize}

In order to give a logical account of the logical relation, we must first define a program logic for the capability machine language. Iris is a separation logic framework, which means we can embed any language into Iris by describing its small step semantics.

While Iris is a framework and supports many languages, it is geared towards models of higher-level languages. Embedding a language asks for a definition of expressions, values, and the state, which abstracts away from the fact that programs are stored in memory.

Typically a machine will have a special register; the program counter, which contains a pointer to an address in memory. That address in turn will contain an integer, which can then be decoded to various machine dependent instructions, such as Load, Store, Jump, etc. Once an instruction has been executed, the program counter is incremented and will now point to the next instruction in memory.

To capture the semantics of the Capability Machine in the Iris framework, we introduce an abstract notion of an instruction, whose operational meaning is to execute the instruction the program counter points to, and an abstract notion of values for halted and failed configurations.

Next we use the Iris framework to prove Hoare triples for each instruction. Like in \cite{Jensen:2013}, we define a notion of points-to for registers. Since a capability machine replaces pointers with capabilities, we will also replace the conventional points-to for memory with a points-to assertion with a permission attached to it: 
$$ r \mapsto_r w \hspace{1cm} a \mapsto_a[p]~w $$
The capability points-to predicate states that address $a$ points to word $w$ with permission $p$. The permission restricts the words an address can update to. For instance, in $a \mapsto_a[\textsc{\tiny RO}]~w$, the $\textsc{\tiny RO}$ permission does not allow for any updates to $a$.

Similar to \cite{Jensen:2013}, the program logic rules for each instruction are of the following form: 
\begin{align*}
\small
	\textit{run time conditions } \wedge &\textit{ decode}(w) = instr \Rightarrow  \\
	\{\{\{ \textsf{PC} \mapsto_r ((p,g),b,e,a) &* a \mapsto_a[p] w * ... \}\}\} \\ 
	\textsf{Instr Executable }\\
	\{\{\{\textsf{PC} \mapsto_r ((p,g),b,e,a&+1) * a \mapsto_a[p] w * ... \}\}\}
\end{align*} 
Where the run time conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is the abstract expression for executing the next instruction in memory. Unlike \cite{Jensen:2013}, the decoding function is in our case assumed. We now have the necessary building blocks for implementing the logical relation.

\section{Logical Relation}

\begin{figure*}[ht]
	\begin{minipage}{\textwidth}
	\begin{align*}
		\mathcal{E}\interp{W}{pc} \triangleq&~\forall r, \mathcal{R}(W)(r) ~*~\textsf{context}(W)(r[\textsf{\small PC}:=pc])\\
		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\{ v, v = HaltedV \implies \exists W' r', W' \sqsubseteq_{priv} W * \textsf{context}(W')(r')\}
	\end{align*}
	\end{minipage}
	\caption{Logical Relation for Expressions}
	\label{fig:expr}
\end{figure*}

\begin{figure*}[ht]
	\begin{minipage}{\textwidth}
	\begin{align*}
		\mathcal{R}\interp{W}{r} \triangleq&~\forall (reg : \text{RegName} \backslash \textsf{PC}), \mathcal{V}\interp{W}{r(reg)}
	\end{align*}
	\end{minipage}
	\caption{Logical Relation for Register States}
	\label{fig:regs}
\end{figure*}

\begin{figure*}[ht]
	\begin{minipage}{\textwidth}
	\begin{align*}
 		\mathcal{V}\interp{W}{z} \triangleq&~\exists z' \in \mathds{Z}. z = z' \\
 		\mathcal{V}\interp{W}{((\text{\tiny{O}},g),b,e,a)} \triangleq&~\top
 \\
 		\left.
		\begin{IEEEeqnarraybox}[\IEEEeqnarraystrutmode \IEEEeqnarraystrutsizeadd{2pt}{2pt}][c]{rCl} 
			\mathcal{V}\interp{W}{((\text{\tiny{RO}},g),b,e,a)} \\
			\mathcal{V}\interp{W}{((\text{\tiny{RW}},g),b,e,a)} \\
			\mathcal{V}\interp{W}{((\text{\tiny{RWL}},g),b,e,a)}
  \end{IEEEeqnarraybox}
  \, \right\}  &\triangleq  \left.\exists~p', p \sqsubseteq p' * \textsf{read\_write\_cond}(\textsf{p',b,e})
  \right. &\textit{where p is {\tiny{RO}}, {\tiny{RW}}, {\tiny{RWL}} resp.}
  \\
   		\left.
		\begin{IEEEeqnarraybox}[\IEEEeqnarraystrutmode \IEEEeqnarraystrutsizeadd{2pt}{2pt}][c]{rCl} 
			\mathcal{V}\interp{W}{((\text{\tiny{RX}},g),b,e,a)} \\
			\mathcal{V}\interp{W}{((\text{\tiny{RWX}},g),b,e,a)} \\
			\mathcal{V}\interp{W}{((\text{\tiny{RWLX}},g),b,e,a)}
  \end{IEEEeqnarraybox}
  \, \right\}  &\triangleq  \left.
	\begin{IEEEeqnarraybox}[
    \IEEEeqnarraystrutmode][c]{rCl}
    \exists~p', p \sqsubseteq p' &*& \textsf{read\_write\_cond}(\textsf{p',b,e})\\
  &*& ~\square~\textsf{exec\_cond(W)(\textsf{p,g,b,e})}
  \end{IEEEeqnarraybox}  
  \right. &\textit{where p is {\tiny{RX}}, {\tiny{RWX}}, {\tiny{RWLX}} resp.}
  \\
 		\mathcal{V}\interp{W}{((\text{\tiny{E}},g),b,e,a)} \triangleq&~\square~\textsf{enter\_cond}(W)(\textsf{g,b,e,a})
	\end{align*}
	\end{minipage}
	\caption{Logical Relation for Words}
	\label{fig:relation}
\end{figure*}

\begin{figure*}[ht]
	\begin{minipage}{\textwidth}
	\begin{align*}
 		\textsf{read\_write\_cond}(p,b,e) \triangleq&~\bigsep_{a \in [b,e]} \text{rel}(a,p,\mathcal{V})\\
 		\textsf{exec\_cond(W)(p,g,b,e)} \triangleq&~\forall a \in [b~e], W' \sqsubseteq^g W.~\rhd~\mathcal{E}\interp{W'}{((p,g),b,e,a)}\\
 		\textsf{enter\_cond(W)(g,b,e,a)} \triangleq&~\forall W' \sqsubseteq^g W.~\rhd~\mathcal{E}\interp{W}{((\textsf{\tiny{RX}},g),b,e,a)}
	\end{align*}
	\end{minipage}
	\\[2em]
	where $W' \sqsubseteq^{\tiny Global} W \iff W' \sqsubseteq_{priv} W$\\ and 
	$W' \sqsubseteq^{\tiny Local} W \iff W' \sqsubseteq_{pub} W$

	\caption{Capability Conditions}
	\label{fig:condition}
\end{figure*}

\subsection{The Value Relation}
%\begin{itemize}
%	\item definition of the relation and World
%	\item LS definitions
%	\item now: standard islands are not used 
%\end{itemize}

The logical relation is a unary relation that relates a word to a notion of \textit{capability safety}. The value relation $\mathcal{V}$ is an Iris relation of type $World \to Word \to iProp$, where the World is a collection of state transition systems used to reason about local state.

This notion of World is different from the logical relation given by \cite{skorstengaardESOP18}, where the World is a collection of invariants describing the behaviour of all memory. Skorstensgaard et. al. use the World to define validity of capabilities, by relating the invariant of the region in question to certain standard islands for capabilities.

@@

We want the value relation to give us the points-to predicate, while asserting that what the predicate points to is valid. The challenge is now to define how that should be done.

\subsection{Reasoning about Local State}
%\begin{itemize}
%	\item How to reason about local state in high level languages
%	\item public/private distinction
%	\item TT style
%	\item difficulties w/ iris invariants 
%\end{itemize}

Logical accounts of relations for high level languages with references have used Iris invariants to define the validity of a reference. In order to reason about local state, it suffices to add a future world relation only in the definition of the expression relation. This approach works in a high level language, where the calls are inherently well-bracketed.

In our case however, calls are not always well-bracketed. \citeauthor{skorstengaardESOP18} distinguishes between well-bracketed and non well-bracketed calls by respectively using public- and private future world relations at the site of the callback.

%Iris invariants do not make such a distinction. Using invariants alone is therefore not sufficient for our purposes. Instead we explicitly define notions of public and private future world relations, and use these notions to distinguish between well-bracketed and non well-bracketed calls.

A callback is an executable capability. This distinction must therefore be done by the value relation. However, if the value relation is parametrized by a world, we have difficulties using Iris invariants. Asserting $\mathcal{V}(W)(w)$ as invariant does not allow for $W$ to change.

The difficulty lies in supporting reasoning about local state, while distinguishing between public- and private future world relations in the value relation, rather than the expression relation. We implement this by using Iris' saved predicates, which allows us to save a predicate by associating it to a unique name. Using that unique name, we are then able to refer to that predicate somewhere else, and state it with the argument that we want. That way, if we save the value relation predicate and associate it to a location, we can later on state that predicate and apply it to the future world in question.

>>>>

In summary, we present Iris formalizations of 
\begin{itemize}
  \item a program logic for reasoning about capability machine programs.
  \item the logical relation from \cite{skorstengaardESOP18}, 
    which captures capability safety and which can be used to reason about examples
    that rely on WBCF and LSE.
\end{itemize}
All of the technical development is mechanized in Coq using the Coq implementation of Iris.

\todo{Alix: I think we need a short discussion (not specifically here) about the fact that we don't really have a theorem that says something like "being in the logical relation implies LSE and WBCF" as that's not really well defined, but we can prove things about specific programs that rely on it}
\todo{Lars: I just tried to reduce the claim above}

@@

\section{A Program Logic for a Capability Machine}

% \begin{itemize}
% \item we define the operational semantics of the CM
% \item we prove basic hoare triples for each of the atomic instructions
%   of the CM (i.e., for load, store, etc.)
% \item non-trivial
%   \begin{itemize}
%   \item while Iris is a framework and supports many languages, it is
%     geared towards models of higher-level languages which abstract from
%     program counters and the fact that programs are stored in memory.
%     -- here our low-level machine model includes registers, in particular
%     the program counter register, and instructions are stored in memory.
%   \item to capture the semantics of the CM in the Iris framework, 
%     we introduce an abstract notion of instruction, whose
%     operational meaning is to execute the next instruction the program counter
%     points to, and abstract notions of values, failed and halted.
%   \item for each machine instruction, we can then prove a hoare triple
%     which expresses the pre- and post-conditions for executing the instruction
%   \end{itemize}
% \end{itemize}

In this section we give a brief overview of how we define a program logic
for reasoning about the capability machine from \cite{skorstengaardESOP18} in the Iris framework.

While Iris is a framework and supports many languages, it is geared towards models of higher-level languages, which abstract from the fact that programs are stored in memory, and hence come equipped with notions of expressions and values, in addition to the program memory.
In contrast, our low-level capability machine model has no notion of expression and values, it just consists of a memory and a register file.
The program counter register contains a pointer to an address in memory.
That address in turn will contain an integer, which can then be decoded to a machine instruction, such as Load, Store, Jump, etc.
Once an instruction has been executed, the program counter is incremented and will then point to the next instruction in memory.

To capture the semantics of the capability machine in the Iris framework, we introduce an abstract notion of an instruction, whose operational meaning is to execute the instruction the program counter points to, and abstract notions of values, for halted and failed configurations.

Next we use the Iris framework to \emph{prove} a Hoare triple for each instruction.
As in \cite{Jensen:2013}, we define a points-to predicate for registers, denoted $r \mapsto_r w$.
Since a capability machine replaces pointers with capabilities, we replace the conventional points-to predicate of separation logic for pointers with a points-to predicate with a permission attached to it, denoted
$a \mapsto_a[p]~w$. 
% \[
%    \hspace{1cm} a \mapsto_a[p]~w.
% \]
This predicate states that address $a$ points to word $w$ with permission $p$.
The permission restricts how the memory may be updated at address $a$.
For instance, if $a \mapsto_a[\textsc{\tiny RO}]~w$, then the $\textsc{\tiny RO}$ (ReadOnly)
permission only gives us permission to read from address $a$, but not to write to it.

% Similar to \cite{Jensen:2013}, the program logic rules for each instruction are of the following form: 
The Hoare triples for basic instructions take the following form
\begin{align*}
\small
	\textit{run time conditions } \wedge &\textit{ decode}(w) = instr \Rightarrow  \\
	\{\{\{ \textsf{PC} \mapsto_r ((p,g),b,e,a) &* a \mapsto_a[p] w * ... \}\}\} \\ 
	\textsf{Instr Executable }\\
	\{\{\{\textsf{PC} \mapsto_r ((p,g),b,e,a&+1) * a \mapsto_a[p] w * ... \}\}\}
\end{align*} 
Here, the run time conditions correspond to the dynamic checks done by the capability machine, and \textsf{Instr Executable} is the abstract expression for executing the next instruction in memory.
This form of Hoare triple is similar to the one used in \cite{Jensen:2013} but
unlike \cite{Jensen:2013}, the decoding function is in our case assumed.

(Here we have only described the format for Hoare triples for individual instructions; we use a trick involving the standard bind rule of Iris to reason about programs consisting of many instructions, but we omit the description of that from this extended abstract.)

% We now have the necessary building blocks for implementing the logical relation.
\section{Logical Relation}

% \begin{figure*}[ht]
% 	\begin{minipage}{\textwidth}
% 	\begin{align*}
% 		\mathcal{E}\interp{W}{pc} \triangleq&~\forall r, \mathcal{R}(W)(r) ~*~\textsf{context}(W)(r[\textsf{\small PC}:=pc])\\
% 		&\sep~\textsc{WP}~\textsf{Seq (Instr Executable)}~\{ v, v = HaltedV \implies \exists W' r', W' \sqsubseteq_{priv} W * \textsf{context}(W')(r')\}
% 	\end{align*}
% 	\end{minipage}
% 	\caption{Logical Relation for Expressions}
% 	\label{fig:expr}
% \end{figure*}

% \begin{figure*}[ht]
% 	\begin{minipage}{\textwidth}
% 	\begin{align*}
% 		\mathcal{R}\interp{W}{r} \triangleq&~\forall (reg : \text{RegName} \backslash \textsf{PC}), \mathcal{V}\interp{W}{r(reg)}
% 	\end{align*}
% 	\end{minipage}
% 	\caption{Logical Relation for Register States}
% 	\label{fig:regs}
% \end{figure*}

% \begin{figure*}[ht]
% 	\begin{minipage}{\textwidth}
% 	\begin{align*}
%  		\mathcal{V}\interp{W}{z} \triangleq&~\exists z' \in \mathds{Z}. z = z' \\
%  		\mathcal{V}\interp{W}{((\text{\tiny{O}},g),b,e,a)} \triangleq&~\top
%  \\
%  		\left.
% 		\begin{IEEEeqnarraybox}[\IEEEeqnarraystrutmode \IEEEeqnarraystrutsizeadd{2pt}{2pt}][c]{rCl} 
% 			\mathcal{V}\interp{W}{((\text{\tiny{RO}},g),b,e,a)} \\
% 			\mathcal{V}\interp{W}{((\text{\tiny{RW}},g),b,e,a)} \\
% 			\mathcal{V}\interp{W}{((\text{\tiny{RWL}},g),b,e,a)}
%   \end{IEEEeqnarraybox}
%   \, \right\}  &\triangleq  \left.\exists~p', p \sqsubseteq p' * \textsf{read\_write\_cond}(\textsf{p',b,e})
%   \right. &\textit{where p is {\tiny{RO}}, {\tiny{RW}}, {\tiny{RWL}} resp.}
%   \\
%    		\left.
% 		\begin{IEEEeqnarraybox}[\IEEEeqnarraystrutmode \IEEEeqnarraystrutsizeadd{2pt}{2pt}][c]{rCl} 
% 			\mathcal{V}\interp{W}{((\text{\tiny{RX}},g),b,e,a)} \\
% 			\mathcal{V}\interp{W}{((\text{\tiny{RWX}},g),b,e,a)} \\
% 			\mathcal{V}\interp{W}{((\text{\tiny{RWLX}},g),b,e,a)}
%   \end{IEEEeqnarraybox}
%   \, \right\}  &\triangleq  \left.
% 	\begin{IEEEeqnarraybox}[
%     \IEEEeqnarraystrutmode][c]{rCl}
%     \exists~p', p \sqsubseteq p' &*& \textsf{read\_write\_cond}(\textsf{p',b,e})\\
%   &*& ~\square~\textsf{exec\_cond(W)(\textsf{p,g,b,e})}
%   \end{IEEEeqnarraybox}  
%   \right. &\textit{where p is {\tiny{RX}}, {\tiny{RWX}}, {\tiny{RWLX}} resp.}
%   \\
%  		\mathcal{V}\interp{W}{((\text{\tiny{E}},g),b,e,a)} \triangleq&~\square~\textsf{enter\_cond}(W)(\textsf{g,b,e,a})
% 	\end{align*}
% 	\end{minipage}
% 	\caption{Logical Relation for Words}
% 	\label{fig:relation}
% \end{figure*}

% \begin{figure*}[ht]
% 	\begin{minipage}{\textwidth}
% 	\begin{align*}
%  		\textsf{read\_write\_cond}(p,b,e) \triangleq&~\bigsep_{a \in [b,e]} \text{rel}(a,p,\mathcal{V})\\
%  		\textsf{exec\_cond(W)(p,g,b,e)} \triangleq&~\forall a \in [b~e], W' \sqsubseteq^g W.~\rhd~\mathcal{E}\interp{W'}{((p,g),b,e,a)}\\
%  		\textsf{enter\_cond(W)(g,b,e,a)} \triangleq&~\forall W' \sqsubseteq^g W.~\rhd~\mathcal{E}\interp{W}{((\textsf{\tiny{RX}},g),b,e,a)}
% 	\end{align*}
% 	\end{minipage}
% 	\\[2em]
% 	where $W' \sqsubseteq^{\tiny Global} W \iff W' \sqsubseteq_{priv} W$\\ and 
% 	$W' \sqsubseteq^{\tiny Local} W \iff W' \sqsubseteq_{pub} W$

% 	\caption{Capability Conditions}
% 	\label{fig:condition}
% \end{figure*}

% \subsection{The Value Relation}
% \begin{itemize}
% 	\item definition of the relation and World
% 	\item LS definitions
% 	\item now: standard islands are not used 
% \end{itemize}

\newcommand{\World}{\mathrm{World}} \newcommand{\Word}{\mathrm{Word}} We now outline how we define a unary logical relation in Iris that captures capability safety.
We define a value relation $\mathcal{V}$ as an Iris relation of type $\World \to \Word \to iProp$, where $\World$ is a collection of state transition systems used to reason about local state.
(It is well-known how state transistion systems can be defined in Iris via Iris's notion of monoids and ghost state.)

This notion of World is different from the logical relation given by \cite{skorstengaardESOP18}, where the World is a collection of invariants describing the behaviour of all memory, not just local state.
We can use a simpler more abstract notion of world because XXX
\todo{Lars: say why succintly, is it just because we use capability points-to predicates ?
  But that could also have been done before\ldots}

\todo{Lars: the following is too hard to read and understand -- I suggest to omit it and just add a succinct description above}
\todo{begin}
Skorstensgaard et.
al.
use the World to define validity of capabilities, by relating the invariant of the region in question to certain standard islands for capabilities.

@@

We want the value relation to give us the points-to predicate, while asserting that what the predicate points to is valid. The challenge is now to define how that should be done.
\todo{end}
% \subsection{Reasoning about Local State}
% \begin{itemize}
% 	\item How to reason about local state in high level languages
% 	\item public/private distinction
% 	\item TT style
% 	\item difficulties w/ iris invariants 
% \end{itemize}

Logical accounts of logical relations for high level languages with references have used Iris invariants to define semantic validity of reference locations. This approach suffices for reasoning about 
examples involving local state encapsulation in languages where all calls are well-bracketed.

In our case, however, calls are not always well-bracketed (since the capability machine includes general jump instructions).
\citeauthor{skorstengaardESOP18} distinguishes between well-bracketed and non well-bracketed calls by using notions of public- and private future worlds.
Iris invariants do not make such a distinction and thus invariants alone are not sufficient for our purposes.
Instead we explicitly define notions of public and private future world relations, and use these notions to distinguish between well-bracketed and non well-bracketed calls.
We use these relations in combination with Iris' saved predicates, which allow us to save a predicate by associating it to a unique name.
Using that unique name, we are then able to refer to the saved predicate somewhere else, and apply it to
an appropriate future world argument.

% A callback is an executable capability.
% This distinction must therefore be done by the value relation.
% However, if the value relation is indexed by a world, we again have difficulties using Iris invariants.
% Asserting $\exists w, \mathcal{V}(W)(w)$ as invariant does not allow for $W$ to change.

% The difficulty is to support reasoning about local state, while distinguishing between public- and private future world relations in the value relation, rather than the expression relation.
% We implement this by using Iris' saved predicates, which allows us to save a predicate by associating it to a unique name.
% Using that unique name, we are then able to refer to that predicate somewhere else, and state it with the argument that we want.
% That way, if we save the value relation predicate and associate it to a location, we can later on state that predicate and apply it to the future world in question.

XXX Add that we prove FTLR and that the LR can be used to prove examples that rely
on WBCF and LSE.
